\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-09 paper/main.tex
illustrative example walkthrough up to line 19, other minor text updates throughout

\newcommand{\comment}[1]{}
% =================================================

@@

\subsection{Software architecture}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.
The \texttt{davos} package consists of two interdependent subpackages (see Fig.~\ref{fig:package-structure}). The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section~\ref{subsec:smuggle}) and onion comment (see Section~\ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section~\ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementations depending on properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython~\cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the correct implementations for the current notebook environment in a single, consistent location. An additional benefit of this design pattern is that it allows maintainers or users to easily extend \texttt{davos} to support new, updated, or custom notebook variants by creating a new \texttt{davos.implementations} module with any necessary tweaks to the existing helper functions.

%The package also includes stub files (\texttt{.pyi} files) for each module for use with a static type checker, as well as a custom GitHub Actions-based CI test suite that uses

@@

\subsection{Software functionalities}
% Present the major functionalities of the software.
\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Importing \texttt{davos} in an IPython notebook enables an additional Python keyword: ``\texttt{smuggle}'' (see Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in \texttt{import} statement to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it automatically, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment.'' An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple format based on the ``type comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, users provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when installing it) may differ from its top-level module name (i.e., the name used when importing it). In such cases, an onion comment can be used to ensure \texttt{davos} installs the proper distribution if the smuggled package can't be found locally:
%Onion comments are useful in such cases for ensuring \texttt{davos} installs the proper distribution if the smuggled package isn't installed locally:
%Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) differs from its top-level module name (i.e., the name used when importing it). Onion comments are useful in such cases for
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references (e.g., Git~\cite{TorvHama05}  branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this allows \texttt{davos} to take certain logical actions when particular arguments are passed. For example, the \texttt{-I}/\newline\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags will all cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable \texttt{davos}'s non-interactive mode (see Section~\ref{subsec:config}), and installing a smuggled package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{dir} to be prepended to the module search path (\texttt{sys.path}), if necessary, so the package can be imported

@@

Line 6 then uses the \texttt{davos.config} object to enable \texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next two packages, \texttt{NumPy}~\cite{HarrEtal20} and \texttt{pandas}. Because these packages rely heavily on custom C data types, loading them may require restarting the notebook kernel if different versions were already imported during the same interpreter session (see Section~\ref{subsec:config}). This is unlikely to happen with \texttt{NumPy}, as it is loaded explicitly only once (by line 7) and is not required by any packages loaded before it (and therefore not implicitly imported while loading them). However, it's possible a future user could edit this notebook to add additional code before line 7, run the notebook's cells out of order, or even configure their IPython kernel to execute custom code on startup (e.g., as Google Colab notebooks do), any of which could potentially import a different, existing version of \texttt{NumPy} before \texttt{davos} smuggles the version specified in the onion comment. Enabling the \texttt{auto\_rerun} option before smuggling \texttt{NumPy} helps to simplify running the notebook in these scenarios, but the main benefit it affords comes when smuggling the \texttt{pandas} library.

The dataset that the user wants to analyze was provided as a \texttt{pandas.Panel} object, serialized with Python's built-in \texttt{pickle} module. The \texttt{pickle} protocol is a popular choice for data persistence in Python, allowing users to save, share, and load arbitrary objects, though with an important caveat: A ``pickled'' object's class definition must be importable from the same module as when it was saved in order to successfully ``unpickle'' (i.e., load and restore) it later. The \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the user's dataset can be loaded only if their notebook uses a \texttt{pandas} version released prior to that change. However, updates to the \texttt{pandas} library since then have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks).
%Recent updates to the \texttt{pandas} library have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks). However, the \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset they are working with can only be loaded by a version released prior to that change.
\texttt{davos} makes it possible to do both within a single notebook, and with the \texttt{auto\_rerun} option, this process can be fully automated. The first time the notebook is run, lines 10--12 will install and load an older \texttt{pandas} version that defines the \texttt{Panel} class, use it to read the dataset from its serialized file format, convert the dataset from a \texttt{Panel} to a multi-index \texttt{DataFrame}, and write its contents to a CSV file. The \texttt{smuggle} statement on line 14 then installs a newer version of \texttt{pandas} (v1.3.5) for use in the analyses themselves. Because an older \texttt{pandas} version was previously loaded (by line 10), and because the package's C extension modules were modified between the old and new versions, switching to the new version requires restarting the notebook kernel. With \texttt{auto\_rerun} enabled, \texttt{davos} takes care of this automatically and will restart the kernel, re-run the notebook up through the current cell, and also re-queue any other cells that the user had queued before the restart was triggered. This second time the code is run (as well as any future times), \texttt{smuggle} statements that have already been executed once before will run (virutlally) as quickly as import statements, since the specified package versions are already installed. Additionally,

>>>>

\newcommand{\comment}[1]{}
%\let\oldtexttt\texttt
%\renewcommand{\texttt}[1]{\mbox{\oldtexttt{#1}}}
% =================================================

@@

\subsection{Software architecture}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.
The \texttt{davos} package consists of two interdependent subpackages (see Fig.~\ref{fig:package-structure}). The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section~\ref{subsec:smuggle}) and onion comment (see Section~\ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section~\ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementations depending on properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython~\cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the proper implementations for the current notebook environment in a single, consistent location. An additional benefit of this design pattern is that it allows both maintainers and users to easily extend \texttt{davos} to support new, updated, or custom notebook variants by creating a new \texttt{davos.implementations} module that defines its own version of each helper function, modified as much or little as necessary.
%The package also includes stub files (\texttt{.pyi} files) for each module for use with a static type checker, as well as a custom GitHub Actions-based CI test suite that uses

@@

\subsection{Software functionalities}
% Present the major functionalities of the software.
\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Functionally, importing \texttt{davos} in an IPython notebook enables an additional Python keyword: ``\texttt{smuggle}'' (see Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in ``\texttt{import}'' to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it automatically, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment.'' An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple format based on the ``type comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, users provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when installing it) may differ from its top-level module name (i.e., the name used when importing it). In such cases, an onion comment can be used to ensure that \texttt{davos} installs the proper distribution if the smuggled package can't be found locally:
%Onion comments are useful in such cases for ensuring \texttt{davos} installs the proper distribution if the smuggled package isn't installed locally:
%Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) differs from its top-level module name (i.e., the name used when importing it). Onion comments are useful in such cases for
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references (e.g., Git~\cite{TorvHama05}  branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this enables \texttt{davos} to take certain logical actions when particular arguments are passed. For example, each of the \mbox{\texttt{-I}/\texttt{--ignore-installed}}, \texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags will cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable \texttt{davos}'s non-interactive mode (see Section~\ref{subsec:config}), and installing a smuggled package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{davos} to prepend \texttt{dir} to the module search path (i.e., \texttt{sys.path}), if necessary, so the package can be imported.

@@

Line 6 then uses the \texttt{davos.config} object to enable \texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next two packages: \texttt{NumPy}~\cite{HarrEtal20} and \texttt{pandas}. Because these libraries rely heavily on custom C data types, loading the particular versions from the onion comments may require restarting the notebook kernel if different versions were already imported during the same interpreter session (see Section~\ref{subsec:config}). This is unlikely to happen with \texttt{NumPy}, as it is loaded explicitly only once (by line 7) and is not required by any packages loaded before it (and therefore not implicitly imported while loading them). However, it's possible a future user could edit this notebook to add additional code before line 7, run the notebook's cells out of order, or even configure their IPython kernel to execute custom code on startup (e.g., as Google Colab notebooks do), any of which could potentially import a different, existing version of \texttt{NumPy} before \texttt{davos} smuggles the version specified in the onion comment. Enabling the \texttt{auto\_rerun} option before smuggling \texttt{NumPy} helps to simplify running the notebook in these scenarios, but the main benefit it affords comes when smuggling the \texttt{pandas} library.

The dataset that the user wants to analyze was provided as a \texttt{pandas.Panel} object, serialized with Python's built-in \texttt{pickle} module. The \texttt{pickle} protocol is a popular choice for persisting data in Python, allowing users to save, share, and load arbitrary objects, though with an important caveat: 
In order to successfully ``unpickle'' (i.e., load and restore) a ``pickled'' object, its class must be defined in and importable from the same module as when it was saved.
%A ``pickled'' object's class definition must be importable from the same module as when it was saved in order to successfully ``unpickle'' (i.e., load and restore) it later. 
The \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset can be loaded only if the user's notebook uses a \texttt{pandas} version released prior to that change. However, more recent updates to the \texttt{pandas} library since then have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats for data structures (including in Jupyter notebooks).
%Recent updates to the \texttt{pandas} library have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks). However, the \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset they are working with can only be loaded by a version released prior to that change.
\texttt{davos} makes it possible to do both within a single notebook, and with the \texttt{auto\_rerun} option, this process can be fully automated. The first time the notebook is run, lines 10--12 will install and load an older version of \texttt{pandas} that defines the \texttt{Panel} class, use it to read the dataset from its serialized file format, convert the dataset from a \texttt{Panel} to a multi-index \texttt{DataFrame}, and write its contents to a CSV file. The \texttt{smuggle} statement on line 14 will then install a newer \texttt{pandas} version (v1.3.5) for use in the analyses themselves. Because an older \texttt{pandas} version was previously loaded (by line 10), and because the package's C extension modules were modified between the old and new versions, switching to the new version will require restarting the notebook kernel. When \texttt{davos} determines this, it will automatically restart the kernel, re-run the notebook up through the current cell, and also re-queue any subsequent cells that were queued before the restart was triggered. This second time the code is run (as well as any future times), lines 10--12 will be skipped since the CSV-formatted dataset has already been created, and the \texttt{smuggle} statements up through line 14 will execute (virtually) as fast as regular \texttt{import} statements since the required package versions have already been installed.
%This second time the code is run (as well as any future times), the \texttt{smuggle} statements that have already been run once will execute (virtually) as fast as \texttt{import} statements since the required package versions are now installed, and lines 10--12 will be skipped altogether since the CSV-formatted dataset has already been created. It's worth noting that one could instead save the converted \texttt{DataFrame} as a pickle file

Line 16 then uses the \texttt{davos.configure()} function to disable the \texttt{auto\_rerun} option and simultaneously enable two other options: \texttt{suppress\_stdout} and \texttt{noninteractive}. With these options enabled, lines 17--18 \texttt{smuggle} \texttt{TensorFlow}~\cite{AbadEtal15}, a powerful end-to-end platform for building and working with machine learning models, and \texttt{UMAP}~\cite{McInEtal18b}, a library that implements a family of related manifold learning techniques. The onion comment in line 18 also specifies that \texttt{UMAP} should be installed with the optional requirements needed for its ``plot'' and ``parametric\_umap'' features. Together, these two packages depend on 36 other unique packages, most of which have dependencies of their own. And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run. Enabling the \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Additionally, installing more than three dozen packages could result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling \texttt{suppress\_stdout} hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run and result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling the \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Enabling the \texttt{suppress\_stdout} option hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.

%a large amount of standard output text generated by the installer program. The \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely.

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run. The \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Additionally, in the process

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run, and generate a large amount of standard output text while doing so. The \texttt{suppress\_stdout} option hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams from the installer program will be shown alongside the regular Python traceback.

% The \texttt{suppress\_stdout} option hides the output from the installer program for these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both its stdout and stderr streams will be shown along with the

% have 36 unique dependencies, most of which require other packages, in turn.

% and \texttt{UMAP}~\cite{McInEtal18b}. \texttt{TensorFlow} provides an end-to-end platform for building and working with machine learning models, and \texttt{UMAP} is a library for non-linear dimensionality reduction. the onion comment in line 18 also specifies that

%These options are enabled while smuggling the following two libraries \texttt{TensorFlow}~\cite{AbadEtal15}, a powerful, heavyweight machine learning

%The next two packages smuggled, \texttt{TensorFlow}~\cite{AbadEtal15} and \texttt{UMAP}~\cite{McInEtal18b} (lines 17--18), are machine learning libraries that depend on a relatively large number of heavyweight packages. The onion comment on line 18 also specifies that \texttt{UMAP} should be installed with

% The following two lines then \texttt{smuggle} \texttt{TensorFlow}~\cite{AbadEtal15}, a large library

Next, the user installs a pair of packages that extend the Jupyter notebook interface with various interactive JavaScript widgets.
